---
title: "Bank Marketing Classification Project"
author: "Atharva Rane"
format: pdf
execute:
  echo: true
  message: false
  warning: false
---

# Introduction

In this project, we analyze the **Bank Marketing** dataset (`bank-additional.csv`) from the UCI Machine Learning Repository.  
The goal is to predict whether a client subscribes to a **term deposit** (`y`: `yes`/`no`) based on client attributes, previous campaign outcomes, and macroeconomic indicators.

We:

- Explore and visualize the data.
- Fit multiple supervised learning models:
  - Logistic regression
  - Penalized logistic regression (LASSO)
  - Classification tree
  - Random forest
- Compare predictive performance using a train–test split.
- Interpret predictor contributions via coefficients and variable importance.
- Assess model stability using bootstrap resampling of the AUC.

---

# Data Loading and Preprocessing

```{r}
library(corrplot)
library(tidyverse)
library(pROC)
library(rpart)
library(rpart.plot)
library(randomForest)
library(glmnet)
library(boot)

# Read data
bank <- read.csv("bank-additional.csv", sep = ";")

# Response as factor with fixed levels
bank$y <- factor(bank$y, levels = c("no", "yes"))

# Drop duration and default (duration is post-call, default is ambiguous/sparse)
bank <- bank %>%
  select(-duration, -default)

# Make categorical variables factors
bank <- bank %>%
  mutate(
    job         = factor(job),
    marital     = factor(marital),
    education   = factor(education),
    housing     = factor(housing),
    loan        = factor(loan),
    contact     = factor(contact),
    month       = factor(month),
    day_of_week = factor(day_of_week),
    poutcome    = factor(poutcome)
  )

# Quick structure and outcome balance
str(bank)
table(bank$y)
prop.table(table(bank$y))
```
## Exploratory Data Analysis

### Class balance

```{r}
table(bank$y)
prop.table(table(bank$y))

ggplot(bank, aes(x = y)) +
  geom_bar(fill = "#2C7FB8", color = "white") +
  ggtitle("Class Balance (Full Dataset)") +
  xlab("Subscription (y)") +
  ylab("Count")
```
## Histogram of Age
```{r}
ggplot(bank, aes(x = age)) +
  geom_histogram(bins = 30, fill = "#31A354", color = "white") +
  ggtitle("Distribution of Age") +
  xlab("Age") +
  ylab("Count")
```
```{r}
ggplot(bank, aes(x = age, fill = y)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 30) +
  ggtitle("Age Distribution by Subscription Outcome") +
  xlab("Age") +
  ylab("Count")
```
## Histogram of Contact Campaign
```{r}
ggplot(bank, aes(x = campaign)) +
  geom_histogram(bins = 25, fill = "#FDB863", color = "white") +
  xlim(0, 20) +
  ggtitle("Distribution of Contact Frequency (campaign)") +
  xlab("Number of contacts during this campaign") +
  ylab("Count")
```
## Preious Outcome vs subscriptons
```{r}
ggplot(bank, aes(x = poutcome, fill = y)) +
  geom_bar(position = "fill") +
  ggtitle("Previous Outcome vs Subscription") +
  ylab("Proportion") +
  xlab("poutcome") +
  scale_y_continuous(labels = scales::percent)
```
## Month vs subscription rate
```{r}
ggplot(bank, aes(x = month, fill = y)) +
  geom_bar(position = "fill") +
  ggtitle("Subscription Rate by Month") +
  ylab("Proportion") +
  xlab("Month") +
  scale_y_continuous(labels = scales::percent)
```
## Euribor rate vs subscription (density plot)
```{r}
ggplot(bank, aes(x = euribor3m, color = y)) +
  geom_density(linewidth = 1.1) +
  ggtitle("Euribor 3-Month Rate by Subscription Outcome") +
  xlab("euribor3m") +
  ylab("Density")
```
### Job vs subscription (proportions)
```{r}
ggplot(bank, aes(x = job, fill = y)) +
  geom_bar(position = "fill") +
  coord_flip() +
  ggtitle("Job Category vs Subscription") +
  ylab("Proportion") +
  xlab("Job") +
  scale_y_continuous(labels = scales::percent)
```
```{r}
# Train–test split (70/30)
set.seed(123)
n <- nrow(bank)
train_idx <- sample(1:n, 0.7 * n)
train <- bank[train_idx, ]
test  <- bank[-train_idx, ]

prop.table(table(train$y))
prop.table(table(test$y))

```
## Logistic Regression baseline model
```{r}
logit_fit <- glm(y ~ ., data = train, family = binomial)
summary(logit_fit)

logit_prob <- predict(logit_fit, test, type = "response")
logit_pred <- ifelse(logit_prob > 0.5, "yes", "no") |> factor(levels=c("no","yes"))

table(logit_pred, test$y)
logit_auc <- auc(roc(test$y, logit_prob))
logit_auc
```
## Penalized Logistic Regression (LASSO)
```{r}
x_train <- model.matrix(y ~ . - 1, train)
y_train <- train$y

set.seed(123)
cv_lasso <- cv.glmnet(x_train, y_train, family="binomial", alpha=1)

best_lambda <- cv_lasso$lambda.min
lasso_fit <- glmnet(x_train, y_train, family="binomial", alpha=1, lambda=best_lambda)

x_test <- model.matrix(y ~ . - 1, test)
lasso_prob <- predict(lasso_fit, newx = x_test, type="response")[,1]
lasso_pred <- ifelse(lasso_prob > 0.5, "yes", "no") |> factor(levels=c("no","yes"))

table(lasso_pred, test$y)
lasso_auc <- auc(roc(test$y, lasso_prob))
lasso_auc

```
## Ridge Regression
```{r}
# Prepare input matrices
x <- model.matrix(y ~ ., data = train)[, -1]   # remove intercept column
y <- train$y

set.seed(123)

# Fit ridge with cross-validation
ridge_cv <- cv.glmnet(
  x,
  y,
  alpha = 0,                # ridge penalty
  family = "binomial",
  type.measure = "auc"      # use AUC for evaluation
)

# Best lambda
ridge_lambda <- ridge_cv$lambda.min
ridge_lambda

# Fit final ridge model at optimal lambda
ridge_fit <- glmnet(
  x,
  y,
  alpha = 0,
  lambda = ridge_lambda,
  family = "binomial"
)

# Predictions on test set
x_test <- model.matrix(y ~ ., data = test)[, -1]
ridge_prob <- predict(ridge_fit, newx = x_test, type = "response")

# Compute AUC
library(pROC)
ridge_auc <- auc(test$y, ridge_prob)
ridge_auc
```
## Classification Tree
```{r}
tree_fit <- rpart(y ~ ., data=train, method="class", 
                  control=rpart.control(cp=0.001))

# Choose best cp
best_cp <- tree_fit$cptable[which.min(tree_fit$cptable[,"xerror"]), "CP"]
tree_pruned <- prune(tree_fit, cp=best_cp)

rpart.plot(tree_pruned, type=2, extra=104)

tree_prob <- predict(tree_pruned, test, type="prob")[, "yes"]
tree_pred <- ifelse(tree_prob > 0.5, "yes", "no") |> factor(levels=c("no","yes"))

table(tree_pred, test$y)
tree_auc <- auc(roc(test$y, tree_prob))
tree_auc

```
## No. of Terminal Nodes:
```{r}
sum(tree_pruned$frame$var == "<leaf>")
```
## Random Forest

```{r}
set.seed(123)
rf_fit <- randomForest(y ~ ., data=train,
                       ntree=500, 
                       mtry=floor(sqrt(ncol(train)-1)),
                       importance=TRUE)

rf_prob <- predict(rf_fit, test, type="prob")[, "yes"]
rf_pred <- predict(rf_fit, test)

table(rf_pred, test$y)
rf_auc <- auc(roc(test$y, rf_prob))
rf_auc

varImpPlot(rf_fit)
```
## Model Performance Comparison
```{r}
perf <- data.frame(
  Model = c("Logistic", "LASSO", "Tree", "Random Forest"),
  AUC = c(logit_auc, lasso_auc, tree_auc, rf_auc)
)
perf

```
## Performance Interpretation
```{r}
exp(coef(logit_fit))
coef(lasso_fit)
importance(rf_fit)
varImpPlot(rf_fit)


```

## Bootstrap for stability (AUC of logistic regression)
```{r}
logit_boot_auc <- function(data, indices) {
  # Bootstrap sample
  d <- data[indices, ]
  
  # Fit logistic regression on bootstrap sample
  fit <- glm(y ~ ., data = d, family = binomial)
  
  # Predict on the same fixed test set to keep comparison fair
  probs <- predict(fit, newdata = test, type = "response")
  
  # Return AUC as a single number
  auc(roc(test$y, probs))
}

set.seed(123)
boot_res <- boot(data = train, statistic = logit_boot_auc, R = 500)

boot_res

# Bootstrap CI for AUC
boot.ci(boot_res, type = "perc")
```

